---
title: "The *DALSM* R-package"
output: github_document
editor_options: 
  chunk_output_type: inline
---

<!-- ```{r setup, include=TRUE}
%knitr::opts_chunk$set(echo = TRUE)
```
-->

The *DALSM* R-package enables to fit a **nonparametric double location-scale model to** right- and interval-**censored data**. 

Consider a vector $(Y,\mathbf{z},\mathbf{x})$ where $Y$ is a univariate
continuous response, $\mathbf{z}$ a $p-$vector of categorical covariates,
and $\mathbf{x}$ a $J-$vector of quantitative covariates.
The response could be subject to right-censoring, in which case one
only observes $(T,\Delta)$, where $T=\min\{Y,C\}$, $\Delta=I(Y\leq C)$
and $C$ denotes the right-censoring value that we shall assume
independent of $Y$ given the covariates. The response could also be
interval-censored, meaning that it is only known to lie within an
interval $(Y^L,Y^U)$.

## Model specification
A location-scale model is considered to describe the distribution of
the response conditionally on the covariates,
$$Y = \mu(\mathbf{z},\mathbf{x}) + \sigma(\mathbf{z},\mathbf{x})\varepsilon$$
where $\mu(\mathbf{z},\mathbf{x})$ denotes the conditional
location, $\sigma(\mathbf{z},\mathbf{x})$ the conditional dispersion,
and $\varepsilon$ an error term independent of $\mathbf{z}$ and
$\mathbf{x}$ with fixed first and second order moments, $\mathrm{E}(\varepsilon)=0$ and
$\mathrm{V}(\varepsilon)=1$.

Assume that independent copies 
$(y_i,\mathbf{z}_ i,\mathbf{x}_ i)~(i=1,\ldots,n)$
are observed on $n$ units with the possibility of
right or interval-censoring on $y_i$  as described above.
Additive models for the conditional location and
dispersion of the response are specified,
$$\mu(z_i, x_i)= \beta_0+ \Sigma_ {k=1}^p \beta_k z_{ik}+\Sigma_ {j=1} ^{J} {f_j ^{\mu}}(x_ {ij})$$
$$\log\sigma(z_i,x_i)= \delta_0+\Sigma_ {k=1}^p \delta_k z_ {ik}+\Sigma_ {j=1}^{J} {{f_{j}^{\sigma}}}(x_ {ij})$$
where $f_{j}^{\mu}(\cdot)$ and $f_{j}^{\sigma}(\cdot)$ denote smooth
additive terms quantifying the effect of the $j {th}$ quantitative
covariate on the conditional mean and dispersion.
The additive terms in the conditional location and dispersion models can be approximated using linear combinations of (recentered)
B-splines,
$$f^ {\mu}_j=\left(\Sigma_ {\ell=1}^L {b_ {j\ell}}(x_ {ij})\theta_ {\ell j}^\mu\right)_ {i=1}^n=\mathbf{B}_ j\pmb{\theta}_j^\mu~~;~~ 
f^ {\sigma}_j=\left(\Sigma_ {\ell=1}^L {b_ {j\ell}}(x_ {ij})\theta_ {\ell j}^\sigma\right)_ {i=1}^n = \mathbf{B}_j\pmb{\theta}_j^\sigma$$
where $[\mathbf{B}_j]_{i\ell}={b_{j\ell}}(x_{ij})$.

Gaussian Markov random field (GMRF) priors penalizing changes in neighbour B-spline parameters are taken,
$$p(\pmb{\theta}^\mu_{j}|\lambda_j^{\mu}) \propto
\exp\left\{-{1\over 2}~{\pmb{\theta}^\mu_{j}}^\top (\lambda^{\mu}_{j}\mathbf{P}^\mu) \pmb{\theta}^\mu_{j} \right\}
~~;~~
p(\pmb{\theta}^\sigma_{j}|\lambda_j^{\sigma}) \propto
\exp\left\{-{1\over 2}~{\pmb{\theta}^\sigma_{j}}^\top (\lambda^{\sigma}_{j}{\mathbf{P}^\sigma}) \pmb{\theta}^\sigma_{j} \right\},
$$

where $\mathbf{P}^\mu, \mathbf{P}^\sigma$ are penalty matrices and
$\lambda^{\mu}_{j},\lambda^{\sigma}_{j}$ positive penalty parameters for the location and dispersion submodels, respectively.
For penalty matrices of order 2, increasingly large values for these parameters point to linear effects for the additive terms.

## Nonparametric error density
The error term $\varepsilon$ is assumed to have fixed first and second order moments, $\mathrm{E}(\varepsilon)=0$ and
$\mathrm{V}(\varepsilon)=1$. To get away from the constraining and sometimes unrealistic normality hypothesis, a flexible form based on P-splines is indirectly assumed for the error density through the log of the hazard function. Lagrange multipliers are used to force mean and variance constraints. We refer to Section 2.5 and Algorithm 3 in [Lambert (2021)](http://doi.org/10.1016/j.csda.2021.107250) for more details and to function **densityIC** in the **DALSM** package for computational aspects to handle right- and interval-censored data.

## Selection of penalty parameters
Marginal posterior distributions of the penalty parameters in the error density and in the location and dispersion submodels are approximated by starting from the following identity,
$$
p(\pmb{\lambda}|{\cal D}) =
{p(\pmb{\psi},\pmb{\lambda}|{\cal D})
\over
p(\pmb{\psi}|\pmb{\lambda},{\cal D})
}
$$
where $\lambda$ and $\pmb{\psi}$ generically denote the penalty and the penalized parameters. Thanks to the GMRF prior for $(\pmb{\psi}|\lambda)$, a Laplace approximation to the conditional posterior in the numerator is relevant. When that expression is evaluated at the conditional posterior mode $\hat{\pmb{\psi}}_\lambda$, one obtains the following approximation to the marginal posterior of the penalty parameters
$$
\tilde{p}(\pmb{\lambda}|{\cal D}) \propto
p(\hat{\pmb{\psi}}_\lambda,\pmb{\lambda}|{\cal D})
~{\begin{vmatrix}\Sigma_\lambda\end{vmatrix}}^{1/2}
$$

where $\Sigma_\lambda$ denotes the conditional variance-covariance matrix. The mode of $\tilde{p}(\pmb{\lambda}|{\cal D})$ could be chosen if point values for the penalty parameters are desired. The combination of Bayesian P-splines with the preceding approximation to the marginal posterior of $\pmb{\lambda}$ is shortly named *Laplace P-splines*.

## Model fitting
Model fitting alternates the estimation of the regression and spline parameters in the location-scale model (see Algorithm 1 in [Lambert 2021](http://doi.org/10.1016/j.csda.2021.107250)), the selection of the penalty parameters (see Algorithm 2) and the update of the error density (see Algorithm 3). The whole procedure is implemented in function **DALSM** of the package.

## The *DALSM* package in action
The data of interest come from the European Social Survey (ESS 2016) with a focus on the money available per person in
Belgian households for respondents aged 25-55 when the main source of
income comes from wages or salaries ($n=756$). Each person reports the
total net monthly income of the household in one of 10 decile-based
intervals: $1: <1\,120$ $(n_1=8)$, $2: [1\,120,1\,399]$ $(n_2=13)$,
$3: [1\,400,1\,719]$ $(n_3=47)$, $4: [1\,720,2\,099]$ $(n_4=53)$,
$5: [2\,100,2\,519]$ $(n_5=82)$, $6: [2\,520,3\,059]$ $(n_6=121)$,
$7: [3\,060,$ $3\,739]$ $(n_7=167)$, $8:[3\,740,4\,529]$ $(n_8=126)$,
$9: [4\,530,5\,579]$ $(n_9=74)$, $10: \geq 5\,580$ euros $(n_{10}=65)$.

We model the relationship between disposable income per person ($91.4\%$
interval-censored, $8.6\%$ right-censored) and the availability of (at
least) two income ($64.2\%$) in the household, as well as the age
($41.0\pm8.83$ years) and number of years of education
completed ($14.9\pm3.34$ years) by the respondent.
This individualized income is obtained by dividing the household one by the
OECD-modified equivalence (Hagenaars1994), as recommended by the Statistical Office of
the European Union (EUROSTAT).

Let us load the package and visualize typical response values:
```{r DALSM1}
## Package installation from R-CRAN
## install.packages("DALSM")
## Alternatively, installation from GitHub:
## install.packages("devtools")
## devtools::install_github("plambertULiege/DALSM")

## Package loading
library(DALSM)

## Data reading
data(DALSM_IncomeData)
resp = DALSM_IncomeData[,1:2]
head(resp,20) ## Visualize first 20 censored income values
summary(DALSM_IncomeData[,3:5]) ## Summary stats for covariates
```
Most reported income values (in thousand euros) are interval-censored (with finite values for the upper bound of the interval), 
some are right-censored (with an *Inf* value for the upper bound of the interval) for respondents belonging to households whose net monthly income exceeds the last decile. 

The nonparametric double location-scale model (with additive *age* and *eduyrs* terms and a fixed effect for the *twoincomes* indicator can be fitted using the *DALSM* function:
```{r DALSM2}
fit = DALSM(y=resp,
           formula1 = ~twoincomes+s(age)+s(eduyrs),
           formula2 = ~twoincomes+s(age)+s(eduyrs),
           data = DALSM_IncomeData)
print(fit)
```
It suggests an average increase of 252 euros (available per person in the household) 
when the respondent and his/her partner are in paid work (conditionally on
*Age* and *Educ*), while the effect on dispersion is not
statistically significant.  
The effects of *Age* and *Educ* on
the conditional mean and dispersion can be visualized on the plots below with the
estimated additive terms.  
```{r DALSM2b}
plot(fit, new.dev=FALSE)
```

The amount of money available
per person in the household tends to decrease with age (see
$f_1^\mu(\text{age})$) between approximately 27 and 37 (most likely
due the arrival of children in the family) and to increase after 40
(probably thanks to wage increase with seniority and the departure of
children). The log-dispersion significantly increases with
*Age*, see $f_1^\sigma(\mathrm{age})$, with an acceleration over 30. However, the dominant effect
comes from the respondent's level of education, with a difference of
about 1\,000 euros (in expected disposable income per person) between
a less educated (6 years) and a highly educated (20 years) respondent,
see $f_2^\mu(\mathrm{eduyrs})$. The effect on dispersion is also large,
see $f_2^\sigma(\mathrm{eduyrs})$, with essentially an important
contrast between less and highly educated respondents, the latter
group showing the largest heterogeneity. The estimated density for the
error term can also be seen, 
with a right-skewed shape clearly distinguishable from the Gaussian
one often implicitly assumed when fitting
location-scale regression models.


## License
**DALSM**: Nonparametric Double Additive Location-Scale Model (DALSM). Copyright (C) 2021-2023 Philippe Lambert

This program is free software: you can redistribute it and/or modify it under the terms of the GNU General Public License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any later version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for more details.

You should have received a copy of the GNU General Public License along with this program. If not, see https://www.gnu.org/licenses/.

## References
[1] Lambert, P. (2021) Fast Bayesian inference using Laplace approximations in nonparametric double additive location-scale models with right- and interval-censored data. Computational Statistics and Data Analysis, 161: 107250. [doi:10.1016/j.csda.2021.107250](http://doi.org/10.1016/j.csda.2021.107250)

[2] Lambert, P. (2021) R-package *DALSM* (Nonparametric **D**ouble **A**dditive **L**ocation-**S**cale **M**odel)- [R-cran](https://CRAN.R-project.org/package=DALSM) ; GitHub: [plambertULiege/DALSM](https://github.com/plambertULiege/DALSM)
